---
title: "Introduction to the Powers of R"
subtitle: "<br/> plus RStudio, R Markdown"
author: "BU Data Science Association"
institute: "Boston University"
date: "2025"
output:
  xaringan::moon_reader:
    yolo: true
    css: [default, metropolis, metropolis-fonts, "styles.css"]
    lib_dir: libs
    df_print: paged
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


# Summary
.my-style[
- R is a statistical programming language; it does only statistics and data science
- Offers a wide range of statistics-related libraries.
- Used very often in academia (particularly social science), government agencies, data analysis roles
- Became a thing in the 1990s
- Works primarily with 2D dataframes, similar to Python's Pandas library
- Creators: Robert Gentleman and Ross Ihaka
]
---

# R vs Python
### R and Python can be used interchangeably for many tasks, but there are some notable differences
.my-style[
- R is 1-indexed (starts counting at 1, most languages start at 0)
- R has a slightly steeper learning curve but is still commonly used by non-CS people (academia, government agencies, etc)
- The difference in speed is not huge; certain tasks are better suited for one of the other but this isn't a big deal
]
---

# R Studio
.my-style[
- IDE (code editor) specifically for R
- Not required to code in R but I have found it worthwhile
- Keeps track of all dataframes, vectors, and even primitive objects (strings, ints, etc) loaded into memory
- Built-in CSV viewer
- Built-in R terminal
]
---

# R Markdown
.my-style[
- Basically Jupyter Notebooks for R, standard way of interacting with R
- Lets you mix between code sections and markdown (text) sections. 
- Running code in a section: either Run -> Run Current Chunk or better is ctrl+shift+c
- Renders an HTML page containing the output of code
- Certain extensions take this a step further, allowing you to make interactive slideshow presentations and Word documents (These slides are rendered from R markdown!)
]
---

# Syntax
.my-style[
- .R extension for base R, .Rmd for R Markdown
- Comments with #
- Asignment: of “=” to assign variables, r uses “<-” (varname <- value)
  - Same as relational algebra
- print()
  - Also: the last item in a codeblock will be printed even without a print()
- DELETING: rm() function deletes an object from memory
	- Useful when you have lots of data loaded
- VIEWING: ls() function lists data loaded into memory
]

---

# **Syntax 2**
## Arrays
.my-style[
Create vectors (python list) by using the c() function with the elements of the vector
]
<code>
vec1 <- c(1,2,3)
vec2 <- c(2,4,6)
</code>

Indexing uses commas: <code>A[1, 2]</code> gets you row 1, column 2 of a dataframe

***Note that indices start at 1 instead of 0***

---
# Functions and Piping
## Functions
<code>functionname <- function(arguments){
Do some stuff
}</code>

## Piping - the “%>%” operator

Piping allows several functions to be ran in a row
**ex:**  <code>round(exp(diff(log(x))), 1) is the same as
x %>% log() %> diff() %>% exp() %>% round(1)</code>

#### Think of pipes as "*This* and *Then*"

---
name: Libraries

# Libraries
```{r}
# import statements
library(tidyverse) 
# Most important one, has a ton of stuff
library(readr) # To read CSV
library(plotly) # Interactive visualizations, python has its own version
library(DT) # to transfer interactive tables to js
library(readxl) # To read excel files
library(wordcloud) # for wordcloud
library(leaflet) # for interactive maps
library(knitr)
```

`library(<library_name>)` is the R equivalent of importing python packages

---

class: left, top
# Reading in Data (& output)

```{r}
# TODO: Import dogs.csv
```

---

class: left, top

# Basic Data Manipulation
```{css, echo = F}
table {
  font-size: 12px;     
}
```



```{r}
# TODO: Create a table with the 25 youngest dogs
```



---

# Simple Data Exploration
### Helps you get a brief idea of what you're looking at

```{r}
# Print summary information about the data
summary(zips_data)
```


# Print/Display Data
.pull-left[
```{r eval=TRUE}
# TODO: Read in zips.xlsx and display a few of its lines
```
]

.pull-right[
```{r eval=TRUE}
# TODO: Add column names to the zip codes (ZipCode and Population)
```
]

### No need for a print statement!

---

# GGplot2 bar graph

```{r fig.dim=c(6,3), out.width="80%"}
# TODO: Create a bar chart with ggplot comparing the total number of dogs of each breed

```


---

# Sorted Plot

```{r fig.dim=c(6,3), out.width="80%"}
# TODO: Increase the amount of data, add tilted x labels, add data labels, sort it in descending order

```

---

# Interactive plot (plotly)

```{r fig.dim=c(6,5), out.width="80%"}
# TODO: Build an interactive equivalent of the plot above using plotly
```



---

# Combining datasets

```{r}
```

---

# `Select` to filter out columns
```{r}
# TODO: TODO: Select only a few important columns (set aside the three useless ones)
simple_dog <- dog_data %>% select('AnimalName', 'AnimalBirthYear', 'AnimalGender', 'BreedName', 'ZipCode')
head(simple_dog,5)

```

---

# `Filter` to filter out rows
### `Arrange` is another way to sort

```{r}
# TODO: Find all the dogs named Fido ("FIDO" by the standards of the dataset) and sort them by descending Birth Year and then Breed Name
```

---

# `Mutate` to create new column

```{r}
# TODO: Create a crude age column by subtracting the dog's birth year from 2025
```

---

# `Count` for counting

```{r}
# TODO: Count the number of male vs female dogs in the dataset
```

- Returns a table with two columns, the thing you're counting, and its value

---

# Word Counting

```{r}
## Word Cloud requires a table with the word and the number of occurrences
# TODO: Count up the number of occurrences of each name, filter out the following common entries that aren't names: "UNKNOWN", "NAME NOT PROVIDED", "NOT"
```

---

# Word Cloud Function
```{r}
# TODO: Build a wordcloud of the most common dog names

```


# Lat/long data
```{r}
lat_long <- read_csv("data/lat_long.csv")
head(lat_long,3)
```


##  Merging
```{r}
# TODO: Merge simple_dog with lat_long to acquire the latitude and longitude of each dog's zip code
```

---

# Challenge: Boston Terrier data trimming
```{r}
# We're going to create a bubble map visualizing Boston Terrier population in NYC
# Ingredients: dog_data, zips_data, lat_long
# Hint: Use filter, select, count, merge (twice), and mutate
# For each zip code, we need the latitude, longitude, and terriers per 10000 people (call this column Density for the sake of the exercise)

```

---

# Maps!

```{r}
# The syntax for making the actual maps is a little insane so we've included the mostly complete code to build the map
mybins <- seq(0, 36, by=(36/9))
mypalette <- colorBin( palette="YlOrRd", domain=<TODO>,
na.color="transparent", bins=mybins)

map <- leaflet(<TODO>) %>% # TODO
  addTiles() %>%
    setView(-74.00, 40.71, zoom = 12) %>%
  addCircleMarkers(~Longitude, ~Latitude,
    fillColor = ~mypalette(<TODO>), fillOpacity = 1, #TODO
    color="white", radius=~(<TODO>/(35/12))+5, stroke=FALSE) #TODO
```


---
# Boston Terrier Density in NYC
```{r echo=FALSE}
map
```
