---
title: "Introduction to the Powers of R"
subtitle: "<br/> plus RStudio, R Markdown"
author: "BU Data Science Association"
institute: "Boston University"
date: "2025"
output:
  xaringan::moon_reader:
    yolo: true
    css: [default, metropolis, metropolis-fonts, "styles.css"]
    lib_dir: libs
    df_print: paged
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


# Summary
.my-style[
- R is a statistical programming language; it does only statistics and data science
- Offers a wide range of statistics-related libraries.
- Used very often in academia (particularly social science), government agencies, data analysis roles
- Became a thing in the 1990s
- Works primarily with 2D dataframes, similar to Python's Pandas library
- Creators: Robert Gentleman and Ross Ihaka
]
---

# R vs Python
### R and Python can be used interchangeably for many tasks, but there are some notable differences
.my-style[
- R is 1-indexed (starts counting at 1, most languages start at 0)
- R has a slightly steeper learning curve but is still commonly used by non-CS people (academia, government agencies, etc)
- The difference in speed is not huge; certain tasks are better suited for one of the other but this isn't a big deal
]
---

# R Studio
.my-style[
- IDE (code editor) specifically for R
- Not required to code in R but I have found it worthwhile
- Keeps track of all dataframes, vectors, and even primitive objects (strings, ints, etc) loaded into memory
- Built-in CSV viewer
- Built-in R terminal
]
---

# R Markdown
.my-style[
- Basically Jupyter Notebooks for R, standard way of interacting with R
- Lets you mix between code sections and markdown (text) sections. 
- Running code in a section: either Run -> Run Current Chunk or better is ctrl+shift+c
- Renders an HTML page containing the output of code
- Certain extensions take this a step further, allowing you to make interactive slideshow presentations and Word documents (These slides are rendered from R markdown!)
]
---

# Syntax
.my-style[
- .R extension for base R, .Rmd for R Markdown
- Comments with #
- Asignment: of “=” to assign variables, r uses “<-” (varname <- value)
  - Same as relational algebra
- print()
  - Also: the last item in a codeblock will be printed even without a print()
- DELETING: rm() function deletes an object from memory
	- Useful when you have lots of data loaded
- VIEWING: ls() function lists data loaded into memory
]

---

# **Syntax 2**
## Arrays
.my-style[
Create vectors (python list) by using the c() function with the elements of the vector
]
<code>
vec1 <- c(1,2,3)
vec2 <- c(2,4,6)
</code>

Indexing uses commas: <code>A[1, 2]</code> gets you row 1, column 2 of a dataframe

***Note that indices start at 1 instead of 0***

---
# Functions and Piping
## Functions
<code>functionname <- function(arguments){
Do some stuff
}</code>

## Piping - the “%>%” operator

Piping allows several functions to be ran in a row
**ex:**  <code>round(exp(diff(log(x))), 1) is the same as
x %>% log() %> diff() %>% exp() %>% round(1)</code>

#### Think of pipes as "*This* and *Then*"

---

name: Libraries

# Libraries
```{r}
# import statements
library(tidyverse) 
# Most important one, has a ton of stuff
library(readr) # To read CSV
library(plotly) # Interactive visualizations, python has its own version
library(DT) # to transfer interactive tables to js
library(readxl) # To read excel files
library(wordcloud) # for wordcloud
library(leaflet) # for interactive maps
library(knitr)
```

`library(<library_name>)` is the R equivalent of importing python packages

---

class: left, top
# Reading in Data (& output)

```{r}
dog_data <- read_csv("data/dogs.csv") #import from csv
# note the assignment operator.
```

---

class: left, top

# Basic Data Manipulation
```{css, echo = F}
table {
  font-size: 12px;     
}
```

```{r}
dog_data <- dog_data[order(dog_data$AnimalName, decreasing=TRUE),]
head(dog_data,25)
```



---


# Print/Display Data
.pull-left[
```{r eval=TRUE}
zips_data <- read_excel("data/zips.xlsx") 
#import from excel
head(zips_data,3)
```
]

.pull-right[
```{r eval=TRUE}
colnames(zips_data) <- c(
"ZipCode", "Population")
head(zips_data,3)
```
]

### No need for a print statement!

---

# GGplot2 bar graph

```{r fig.dim=c(6,3), out.width="80%"}
head_data <- head(dog_data, 10)
plot <- ggplot(data = head_data, aes(x = BreedName)) + geom_bar(fill="darkgreen")
plot
```


---

# Sorted Plot

```{r fig.dim=c(6,3), out.width="80%"}
head_data <- head(dog_data, 100)
plot1 <- ggplot(head_data, aes(fct_infreq(BreedName))) + geom_bar(fill="maroon") +   theme(axis.text.x = element_text(angle = 45, hjust=1)) + geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.1)
plot1
```

---

# Interactive plot (plotly)

```{r fig.dim=c(6,5), out.width="80%"}
ggplotly(plot1)
```

---

# Simple Data Exploration
### Helps you get a brief idea of what you're looking at

```{r}
summary(zips_data)
```

---

# Combining datasets

```{r}
merged_dog <- merge(dog_data, zips_data, by="ZipCode")
head(merged_dog,5)
```

---

# `Select` to filter out columns
```{r}
merged_dog <- merged_dog %>% select('AnimalName', 'AnimalBirthYear', 'AnimalGender', 'BreedName', 'ZipCode', 'Population')
head(merged_dog,5)

```

---

# `Filter` to filter out rows
### `Arrange` is another way to sort

```{r}
fidos <- merged_dog %>% filter(AnimalName == "FIDO") %>% arrange(desc(AnimalBirthYear), desc(BreedName))
head(fidos,5)
```

---

# `Mutate` to create new column

```{r}
fidos <- fidos %>% mutate(age = 2023 - AnimalBirthYear)
head(fidos,5)
```

---

# `Count` for counting

```{r}
merged_dog %>% count(AnimalGender)
```

- Returns a table with two columns, the thing you're counting, and its value



---

# Word Counting

```{r}
words <- dog_data %>% select(AnimalName) %>% count(AnimalName, sort=TRUE)
words2 <- words %>% filter(AnimalName != "UNKNOWN" & AnimalName != "NAME NOT PROVIDED" & AnimalName != "NOT")
head(words2, 5)
```

---

# Word Cloud Function
```{r}
cloud <- wordcloud(words = words2$AnimalName, freq = words2$n, min.freq = 100,           max.words=200, random.order=FALSE)
cloud
```


# Lat/long data
```{r}
lat_long <- read_csv("data/lat_long.csv")
head(lat_long,3)
```


##  Merging
```{r}
merged_dog <- merge(merged_dog, lat_long, by="ZipCode")
head(merged_dog,5)
```

---

# Boston Terrier data trimming
```{r}
# We're going to create a bubble map visualizing Boston Terrier population in NYC
# Ingredients: dog_data, zips_data, lat_long
# Hint: Use filter, select, count, merge (twice), and mutate
# For each zip code, we need the latitude, longitude, and terriers per 10000 people (call this column Density for the sake of the exercise)
terriers <- dog_data %>% filter(
BreedName == "Boston Terrier")

counter <- terriers %>% count(ZipCode, sort=TRUE)

counter <- merge(counter, zips_data, by="ZipCode")
counter <- merge(counter, lat_long, by="ZipCode")
counter <- counter %>% mutate(Density = 10000 * n/Population) %>% # Calculate a density column representing Boston Terriers per 10000 people in each zip code
select(Density, Latitude, Longitude)
head(counter,5)

# fidos %>% count(ZipCode, sort=TRUE)

```

---

# Maps!

```{r}
mybins <- seq(0, 36, by=(36/9))
mypalette <- colorBin( palette="YlOrRd", domain=counter$Density,
na.color="transparent", bins=mybins)

map <- leaflet(counter) %>% # TODO
  addTiles() %>%
    setView(-74.00, 40.71, zoom = 12) %>%
  addCircleMarkers(~Longitude, ~Latitude, 
    fillColor = ~mypalette(Density), fillOpacity = 1,
    color="white", radius=~(Density/(35/12))+5, stroke=FALSE)
```


---
# Boston Terrier Density in NYC
```{r echo=FALSE}
map
```
